Comparison of simulation to theory

### 2016-02-05 sims/compare-to-theory.Rmd
more on comparison to theory
comparison to theory; changed "F" to "FALSE"

Here are some functions we'll use to extract what we want
from data in the `chunks` format:
```{r helper_fns}
genotype.counts <- function (ch) { 
    gcounts <- t(sapply( ch, function (x) { tabulate( 1+rowSums( x>0 ), nbins=3 ) } )) 
    colnames(gcounts) <- c("AA","AB","BB")
    return(gcounts)
}
dip.to.hap <- rbind( A=c(AA=2,AB=1,BB=0), B=c(AA=0,AB=1,BB=2) )
```

@@

To get the value predicted from theory, we'll solve the corresponding PDE:
```{r selected_cline_theory}
tt <- seq(0,theory.params$tau,length.out=21)
xgrid <- setup.grid.1D(x.up=-theory.params$width/2, x.down=theory.params$width/2, N=2*theory.params$width)
fgrid <- extend_grid(xgrid)
fwds.soln <- forwards_pde(s=theory.params$s, times=tt, grid=fgrid, sigma=theory.params$sigma )
```
... and, compare it to the observed values:
```{r selected_cline_sim}
gcounts <- genotype.counts(chunks[[1]])
# sim frequencies
sfreqs <- sweep( tcrossprod(gcounts,dip.to.hap), 1, 2*rowSums(gcounts), "/" )
# theory frequencies
tfreqs <- cbind( A=fwds.soln[nrow(fwds.soln),-1], B=1-fwds.soln[nrow(fwds.soln),-1] )
matplot(xx, sfreqs, type='l', lty=1, main="Selected allele frequency", ylab="frequency", xlab='geographic position')
matlines(fgrid$x.mid, tfreqs, lty=4, lwd=2)
legend("topright", legend=outer(colnames(sfreqs),c("sim","theory"),paste,sep=" : "), lty=rep(1:2,each=2), col=rep(1:ncol(sfreqs),2))
```

@@

>>>>

Here are some functions we'll use to extract what we want
from data in the `chunks` format:
```{r helper_fns}
genotype_counts <- function (ch) { 
    gcounts <- t(sapply( ch, function (x) { tabulate( 1+rowSums( x>0 ), nbins=3 ) } )) 
    colnames(gcounts) <- c("AA","AB","BB")
    return(gcounts)
}
conditioned_genotype_counts <- function (x,y) { 
    # returns joint genotype counts across space from two chunks objects
    #  a (ndemes) x 4 x 4 array (4 phased genotypes)
    deme <- rep(seq_along(x),sapply(x,NROW))
    x.alleles <- ifelse(do.call(rbind,x)==0,"A","B")
    y.alleles <- ifelse(do.call(rbind,y)==0,"A","B")
    x.genotypes <- factor( paste0(x.alleles[,1],x.alleles[,2]), levels=c("AA","AB","BA","BB") )
    y.genotypes <- factor( paste0(y.alleles[,1],y.alleles[,2]), levels=c("AA","AB","BA","BB") )
    # x.genotypes <- factor( c("AA","AB","BB")[1+rowSums( do.call(rbind, x)>0 )], levels=c("AA","AB","BB") )
    # y.genotypes <- factor( c("AA","AB","BB")[1+rowSums( do.call(rbind, y)>0 )], levels=c("AA","AB","BB") )
    return( table( deme, x=x.genotypes, y=y.genotypes ) )
}
dip.to.hap <- rbind( A=c(AA=2,AB=1,BB=0), B=c(AA=0,AB=1,BB=2) )
bidip.to.bihap <- rbind(
            "A-A" = c( 2, 1, 1, 0, 1, 1, 0, 0, 1, 0, 1, 0, 0, 0, 0, 0 ),
            "A-B" = c( 0, 0, 0, 0, 1, 0, 1, 0, 1, 1, 0, 0, 2, 1, 1, 0 ),
            "B-A" = c( 0, 1, 1, 2, 0, 0, 1, 1, 0, 1, 0, 1, 0, 0, 0, 0 ),
            "B-B" = c( 0, 0, 0, 0, 0, 1, 0, 1, 0, 0, 1, 1, 0, 1, 1, 2 ) )
colnames( bidip.to.bihap ) <- as.vector( outer( c("AA","AB","BA","BB"), c("AA","AB","BA","BB"), paste, sep="-" ) )
conditional_haplotype_counts <- function (gcounts) {
    # uses output of conditioned_genotype_counts to get conditional haplotype counts
    # a (ndemes) x 2 x 2 array
    dim(gcounts) <- c(dim(gcounts)[1],prod(dim(gcounts)[2:3]))
    out <- tcrossprod( gcounts, bidip.to.bihap )
    dim(out) <- c(dim(gcounts)[1],2,2)
    dimnames(out) <- list( dimnames(gcounts)[[1]], c("A","B"), c("A","B") )
    return(out)
}
conditional_freqs <- function (gcounts) {
    # uses output of conditioned_genotype_counts to get conditional allele frequencies
    # a (ndemes) x 2 matrix
    hcounts <- conditional_haplotype_counts(gcounts)
    cbind( A = hcounts[,"A","A"]/(hcounts[,"A","A"]+hcounts[,"A","B"]),
           B = hcounts[,"A","B"]/(hcounts[,"A","B"]+hcounts[,"B","B"]) )
}
```

@@

To get the value predicted from theory, we'll solve the corresponding PDE:
```{r selected_cline_theory}
tt <- seq(0,theory.params$tau,length.out=21)
xgrid <- setup.grid.1D(x.up=-theory.params$width/2, x.down=theory.params$width/2, N=2*theory.params$width)
fgrid <- extend_grid(xgrid)
fwds.soln <- forwards_pde(s=theory.params$s, times=tt, grid=fgrid, sigma=theory.params$sigma )
```
... and, compare it to the observed values:
```{r selected_cline_sim}
gcounts <- genotype_counts(chunks[[1]])
# sim frequencies
sfreqs <- sweep( tcrossprod(gcounts,dip.to.hap), 1, 2*rowSums(gcounts), "/" )
# theory frequencies
tfreqs <- cbind( A=fwds.soln[nrow(fwds.soln),-1], B=1-fwds.soln[nrow(fwds.soln),-1] )
matplot(xx, sfreqs, type='l', lty=1, main="Selected allele frequency", ylab="frequency", xlab='geographic position')
matlines(fgrid$x.mid, tfreqs, lty=2, lwd=2)
legend("topright", legend=outer(colnames(sfreqs),c("sim","theory"),paste,sep=" : "), lty=rep(1:2,each=2), col=rep(1:ncol(sfreqs),2))
```

@@

**Question:** Is that shift due to stochasticity, or have I got the center in the wrong place on the sims?

# Ancestry probabilities

First, compute theoretical predictions at the values of $r$ in the simulation:
```{r linked_soln}
linked.solns <- lapply( sim.rvals, function (r) {
        forwards_backwards_pde(s=theory.params$s, times=tt, grid=xgrid, r=r, 
                               sigma=theory.params$sigma, fwds.soln=fwds.soln, log.p=TRUE)
    } )
```
... and compare to the observed values:
```{r compare_linked}
for (kr in seq_along(sim.rvals)[-1]) {
    gcounts <- conditioned_genotype_counts(chunks[[kr]], chunks[[1]])
    # frequencies from simulation
    cfreqs <- conditional_freqs(gcounts)
    # frequencies from theory
    tfreqs <- linked.solns[[kr]][nrow(linked.solns[[kr]]),-1]
    dim(tfreqs) <- c( length(xgrid$x.mid), 2 )
    matplot(xx, cfreqs, type='l', lty=1, main=sprintf("Frequency of A ancestry by selected background, r=%f",sim.rvals[rk]), ylab="frequency", xlab='geographic position')
    matlines(xgrid$x.mid, tfreqs, lty=2, lwd=2)
    legend("topright", legend=outer(colnames(cfreqs),c("sim","theory"),paste,sep=" : "), lty=rep(1:2,each=2), col=rep(1:ncol(sfreqs),2))
}
```
